\begin{acknowledgments}

  This thesis marks a particularly significant point in a long personal,
  academic, and intellectual journey. There have been so many people who have
  helped me along the way---many in ways I'm likely unaware of at this
  point---that the task of properly acknowledging them is a little overwhelming.
  
  I first want to thank the members of my committee: Florian Jaeger, Dick Aslin,
  Lori Holt, and Jeff Runner. Florian in particular has pushed me to take the
  ideas I've explored here much further than I thought they could go. I can't
  think of anyone with more enthusiasm for ideas, intellectual honesty, and
  broad curiosity.

  The Rochester BCS community provided me a better intellectual home than I'd
  have imagined was possible. The kind of cross-disciplinary inquiry that
  first attracted me to cognitive science is embodied here in a truly special
  way. I particularly want to thank Robbie Jacobs for providing invaluable input
  on my projects at many points, despite the fact that I am a ``language
  person'' and he is a ``vision person''. I also want to thank Rajeev Raizada,
  who taught me everything I know about functional neuroimaging, and
  communicating my work in terms of clear, concrete metaphors.

  Just as importantly, I would never have arrived at Rochester in the first
  place without the guidance of the many people who have mentored me over the
  years. During my fellowship at UMD, Bill Idsardi encouraged my curiosity and
  recognized, in a way that I didn't, that there's not as big a gulf between
  linguistics, Bayesian nonparametrics, and computational neuroscience as there
  might seem.

  Further back, the Williams cognitive science community drew me into this field
  in the first place. Andrea Danyluk and Joe Cruz's captivating seminar
  introduced me to the combination of rigorous computational thinking applied to
  philosophically interesting questions that are grounded in human behavior that
  to this day exemplifies the intellectual potential of the field of cognitive
  science. Safa Zaki took a chance on an enthusiastic frosh who knew absolutely
  nothing about cognitive psychology as a summer research intern. Safa taught me
  that it's okay to mix behavioral and computational approaches. Moreover,
  without Safa's firm encouragement I doubt I would have ended up in grad school
  at all, and I consider myself deeply fortunate to count her as a friend and
  mentor.

  Before that, a great many others have guided, supported, and mentored me. In
  the interest of space, I'll only thank Susan Thibodeau for telling me to shut
  up and listen more. A tough pill to swallow at the time, but it's advice that
  has served me extremely well.

  None of this work would have been possible without the staff who keeps things
  running. Thanks to Kathy Corser, Jen Gillis, Chris Dambra, and Chris
  Freemesser in the BCS department, and to Andrew Watts in the HLP Lab. Thanks
  also to Judy Ripton, Pat Weber, and Evi Vanoost at the RCBI, and Donias Doko
  in the Raizada Lab.

  Finally, my friends and family have had a role in this work in ways that are
  both tangible and intangible. Thanks, in no particular order, to Alex, Judith,
  Masha, Esteban, Dan, Ilker, Amanda, Lauren, Ruth, Liz, Greg, Lizz, Patty,
  Davis, Kevin, Tal, Keturah, Sarah, Irene, Cory, Klinton, Bozena, Darcy, Katie,
  Travis, Elika, Zach, Jue, Liz, Mom, Bob, Steph, Jeff, Eric, and Alyssa. I'm
  sure I've forgotten someone, please forgive me and chalk it up to sleep
  deprivation.

  Thanks, finally, to everyone at Joe Bean, Pour (Glen Edith), and Ugly Duck for
  keeping me sufficiently caffeinated, and to Morgan for running regular therapy
  sessions for years.
  
\end{acknowledgments}
